---
title: "Homework 6"
author: "Daniel Blumenkranz"
date: "10/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(gapminder)
```

## Exercise 1
A defensive function that calculates GDP:
```{r}
gapminder_gdp <- function(#default is all countries and years
                          countries = gapminder$country,
                          years = gapminder$year) {
  # check input
  stopifnot(countries %in% gapminder$country, years %in% gapminder$year)
  
  # calculate gdp for given countries and years
  gapminder %>% 
    filter(country %in% countries & year %in% years) %>% 
    mutate(gdp = gdpPercap*pop) %>% 
    select(country, year, gdp) %>% 
    return()
}
```

Testing the function:
```{r, error=TRUE}
# default arguments
gapminder_gdp()

# Specific country and years
years <- c(1967, 1977, 1987, 1997, 2007, 2017)
gapminder_gdp(countries = "Denmark", years = years)
```

This function is defensive because it checks that the input to the function is what it expects. If not, as in this case, it returns an informative error message. Here, the function tells us that some of the input years are not in the gapminder data set. Let's check which one:
```{r}
years %in% gapminder$year
```
The gapminder data set does not include 2017. Let's remove that and call the function again:
```{r}
years <- years[-6]
gapminder_gdp(countries = "Denmark", years = years)
```

## Exercise 2
How I understand the exercise: Write a for-loop that returns life expectancy info for every year only for countries that start with B.
```{r}
for (i in 1:nrow(gapminder)){
  
  # If the country starts with B the grepl-function will evaluate to TRUE and the following code will be read 
  if (grepl("^B", gapminder$country[i])){
    
    # Below 50
    if (gapminder$lifeExp[i]<50){
      info <- ': Life expectancy was below 50'}
    
    # Between 50 and 70
    else if (gapminder$lifeExp[i]>50 & gapminder$lifeExp[i]<70){
      info <- ': Life expectancy was above 50 and below 70'}
    
    # Above 70
    else {info <- ': Life expectancy was above 70'}
  
  # Output
  print(paste0(gapminder$country[i], ' ', gapminder$year[i], info))
  }
}
```


